A Simple Minibatch Acceptance Test for MCMC

### 2016-05-18 nips_2016.tex
Updates to Section 2

In the standard MCMC method~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the target
distribution $p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated)
samples $\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.
Each iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This can be implemented by drawing
a uniform random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. The $P_a$ from Equation~\ref{eq:traditional}
satisfies detailed balance, so if one samples long enough, one will arrive at a stationary
distribution matching the posterior, though a burn-in period and/or taking every $n$th sample may be
helpful.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the mini-batch size in one iteration; in the worst case, all $N$ data points may be needed in a
single iteration.

A similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical
properties, but comes at the cost of more computational time. The work
of~\cite{conf/uai/MaclaurinA14} presents another minibatch MCMC method, but which gets the
\emph{exact} posterior, with the only requirement being that there needs to be a lower bound on the
log likelihood for each data point. They augment the likelihood with auxiliary random variables, but
this can hurt mixing.

There is a separate line of work related to optimizing MCMC about simulating the physics of a
probability distribution. By viewing random variables as ``particles'' in a system, it is possible
to apply Hamiltonian Monte Carlo (HMC)~\cite{mcmc_hamiltonian_2010} methods which generate both high
quality and distant proposals, getting the best of both worlds. These require a full gradient
computation of the posterior, which means they also suffer from the problem of needing to know the
likelihood at each point. Stochastic versions of HMC~\cite{sghmc_2014,stochastic_thermostats_2014}
use a minibatch of data each iteration but require controlling the noise from sampling so the sample
sequence does not diverge. A similar approach is to use Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12}, which does not use the momentum terms in HMC.

Minibatch SGHMC and Langevin Dynamics are orthogonal to our work; two ways of attempting
to solve the same problem of MCMC methods with big datasets. We could augment their methods with our
MH test, to get the benefit of their high quality proposals. Note that while their methods in theory
do not need an MH test, in practice, they need one due to discretization of the system. Furthermore,
tests may also be needed for when the algorithms are run with a constant step size. {\color{blue}
Daniel: not totally confident on this paragraph, how much SGHMC/LD do we discuss?}

>>>>

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. The $P_a$ from Equation~\ref{eq:traditional}
satisfies detailed balance, so if one samples long enough, one will arrive at a stationary
distribution matching the posterior, though a burn-in period and/or taking every $n$th sample may be
helpful.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the mini-batch size; in the worst case, all $N$ data points may be needed in a single iteration.  A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties,
but comes at the cost of more computational time.

The work of~\cite{conf/uai/MaclaurinA14} presents another minibatch MCMC method which gets the
\emph{exact} posterior, so long as there exists a (cheap) lower bound on the log likelihood for each
data point. Their approach is only a starting point, however, and it can be difficult to derive
these lower bounds in general cases. The extra auxiliary variables they introduce also hurt mixing.

There is a separate line of MCMC work on simulating the physics of a probability distribution. By
viewing random variables as particles in a system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate the best case scenario for proposals: high
quality \emph{and} distant.  HMC methods require a full gradient computation of the posterior, which
means they also face the problem of computing the likelihood at each point. Stochastic versions of
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use a minibatch of data each iteration but require
controlling the noise from sampling to avoid divergence. Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a similar approach which does not use the
momentum terms in HMC.  Minibatch SGHMC and Langevin Dynamics are orthogonal to our work and are
other ways of attempting to solve the same problem of MCMC methods on big data. We can combine our
methods with these results, as we demonstrate in Section~\ref{ssec:nets}.
